General purpose gene annotation using deep learning

### 2020-04-07 paper/main.tex
few things here and there

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to effectively blank out those bases during training. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{Tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes can be found in Table \ref{Tab:statistics_training}. For the training, we divided each continous genomic sequence into 20.000 bases long subsequences, which are then used as input or label for the neural network. We appended zero padding if the subsequences are not long enough. If a subsequence is fully marked as erroneous, we do not include it in the training or evaluation datasets.

@@

\subsection{Model Architecture and Usage}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

@@

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be any of 20.000, 50.000, 100.000 or 200.000 and is determined by the species family(?). Families that tend to have longer genes get assigned longer sequences. If the N75 of a species is less than twice as high as the supposed sequence length it is lowered until either this criteria is met or a length of 50.000 is reached. For implementation details on this and the following techniques see Supplementary Material (?) or the source code..

We also construct the final predictions of a single model by overlapping predictions that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. Supplementary Material (?) contains the influence on overlapping with respect to the position in a sequence for each genome, ordered by N75. We found that overlapping works best if the genomes are not very fragmented.

@@

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{Tab:03}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Augustus & 0.632 (+-0.089) & 0.749 (+-0.161)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup.}
\end{table}

\section{Results}

We evaluate the performance of our final model base on ... The overall results are shown in table \ref{Tab:03}.

@@

\section{Supplementary Material}
\subsection{Dilated CNN Baseline Network Search}

@@

\subsection{Training Data statistics}
\begin{table}[!t]
\processtable{Data group statistics for our training data\label{Tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{The description of Table \ref{Tab:statistics_all} applies here as well.}
\end{table}

>>>>

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation (explanation why this is a sign of error?). As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to effectively blank out those bases during training. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{Tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

For the training, we divided each continous genomic sequence into 20.000 bases long subsequences, which are then used as input or label for the neural network. We appended zero padding if the subsequences are not long enough. If a subsequence is fully marked as erroneous, we do not include it anywhere.

@@

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

@@

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be any of 20.000, 50.000, 100.000 or 200.000 and is determined by the species family(?). Families that tend to have longer genes get assigned longer sequences. For more implementation details on this see Section \ref{sec:longer} or the source code.

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. Supplementary Material (?) contains the influence on overlapping with respect to the position in a sequence for each genome, ordered by N75. We found that overlapping works best if the genomes are not very fragmented.

@@

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{Tab:03}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Augustus & 0.632 (+-0.089) & 0.749 (+-0.161)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup. The Augustus result is the median score for all genomes.}
\end{table}

\section{Results}
We evaluate our performance based on the average cross-species Subgenic F1 for all genomes except the few used during training. The results are displayed in Table \ref{Tab:03} and show a clear improvement over the AUGUSTUS and dilated CNN baselines. We also demonstrate the effectiveness of two techniques used during inference with the usage of longer input lengths being enabled by our LSTM stack without any fully connected layers on top. To our knowledge, this technique was not used before with a model used for gene annotation.

@@

\section{Supplementary Material}
\subsection{Dilated CNN Baseline Network Search}
\label{sec:dcnn}

@@

\subsection{Training Data statistics}
\label{sec:training_data}

\begin{table}[!t]
\processtable{Data group statistics for our training data\label{Tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{The description of Table \ref{Tab:statistics_all} applies here as well.}
\end{table}

\subsection{Longer Sequence Input}
\label{sec:longer}
Implementation detail:
If the N75 of a species is less than twice as high as the supposed sequence length it is lowered until either this criteria is met or a length of 50.000 is reached.

Table of species with their length.
